A Shell-like Model for General Purpose Programming

### 2018-12-07 doc/paper.tex
paper updates

% \documentclass[format=sigconf, review=true, draft=true, screen=true]{acmart}
\documentclass[sigplan,10pt,nosumlimits]{acmart}

@@

\usepackage{listings}

\title{Pipe-based Programming}
\author{Jeanine Adkisson}
\affiliation{\institution{Tokyo Institute of Technology}}
\email{jneen@prg.is.titech.ac.jp}

@@

\begin{document}

@@

@@

We explore the implementation and design challenges that come with such a model, and also show that this model is able to easily create and compose concurrent algorithms.
The UNIX shell programming model has been shown to be a successful model for composing programs - spawning unrelated programs in parallel to communicate over operating-system pipes.

@@

Since UNIX pipes can only manage bytes, usually channels with a fixed buffer are specified for performance reasons. However, in the languages we consider, we assume we can transactionally write a single complex value to a channel, resulting in the ability to send data in batch without resorting to a queue that may interleave writes.

A core requirement of pipe-based languages is a semantic of process \emph{interruption}: automatic clean-up of processes that will no longer be used. Consider the following pseudo-code for a pipe-based language:

\begin{verbatim}
  read-lines tmp/large-file
    | take 10
    | each (?line => do-expensive-work $line)
\end{verbatim}
In this example, three processes are spawned: (1) a process with a file open that writes one line at a time to its output, (2) a process that reads 10 times from its input, writes each entry to the output, and then exits, and (3) a process that reads every input and calls a function to perform an expensive task.
A user's intent when typing such code may be to read 10 lines from a file and synchronously perform an action on each line.

@@

\section{Design Considerations}
\paragraph{Compensation.} In order to observe interruption, we define a system of \emph{compensation actions}, which will be run in the case of an interruption.
For example, the function \verb/file-lines/ above could be implemented as:

\begin{verbatim}
(file-lines ?fname) = (
  handle = (open-file $fname) %%! close-file $fname
  until (=> eof? $handle) (=> read-until "\n" $handle)
  checkpoint
)
\end{verbatim}

In this code, the \verb/%%!/ symbol registers a compensation action that will be run either at the next \verb/checkpoint/ invocation, or when the process is interrupted. In this way, we can ensure that the file is closed when the process exits. Compensations are the only way in which interruptions can be directly observed by client code - without compensations, the only way of observing forever-blocked processes would be to inspect the process table, or observe the memory footprint of the program. Thus, the primary task of our system is \emph{to run a process's compensation actions at the appropriate time}.

@@

Consider the following example, with the assumption that \verb/do-other-operation/ does not write to the standard output:

@@

\paragraph{Channel registration.}

@@

@@

\section{Description of Semantics.}

>>>>

\documentclass[format=sigconf, review=true, draft=true, screen=true]{acmart}
% \documentclass[sigplan,authorversion,draft]{acmart}

@@

\usepackage{listings}
\lstset{
language=bash,
% backgroundcolor=\color[rgb]{0.95, 0.95, 0.95},
tabsize=2,
rulecolor=,
basicstyle=\ttfamily,
upquote=true,
% aboveskip={1.5\baselineskip},
columns=fullflexible,
% columns=fixed,
showstringspaces=false,
extendedchars=true,
breaklines=true,
prebreak = \raisebox{0ex}[0ex][0ex]{\ensuremath{\hookleftarrow}},
frame=L,
showtabs=false,
showspaces=false,
showstringspaces=false,
% identifierstyle=\ttfamily,
keywordstyle=\color[rgb]{1.0,0,0}\bfseries,
keywordstyle=[1]\color[rgb]{0,0,0.75}\bfseries,
keywordstyle=[2]\color[rgb]{0.5,0.0,0.0},
keywordstyle=[3]\color[rgb]{0.127,0.427,0.514},
keywordstyle=[4]\color[rgb]{0.4,0.4,0.4},
commentstyle=\color[rgb]{0.5,0.5,0.5}\itshape,
stringstyle=\color[rgb]{0.639,0.082,0.082},
morekeywords={read-lines, take, each},
deletekeywords={do, read},
%numbers=left,%
numbersep=5pt,%
numberstyle=\tiny\color{gray},%
emphstyle=\bfseries,%
breaklines=true,
breakatwhitespace=true,%
escapechar=`,
%numbers=left,
}
\lstset{escapeinside={<@}{@>}}
\definecolor{ttblue}{rgb}{0,0,0.75}
\title{A Shell-like Model for General Purpose Programming}
\author{Jeanine Adkisson}
\affiliation{\institution{Tokyo Institute of Technology}}
\email{jneen@prg.is.titech.ac.jp}

\author{Johannes Westlund}
\affiliation{\institution{TODO}}
\email{TODO}
\begin{document}

@@

@@

% TODO LIST
% * Find an example of named pipes reopening
% * Find an example of a non-linear pipe arrangement (prob. from flow-based programming)
% * Plan the Description of Magritte section
% * Add Call Semantics to Design Considerations
% * Institute and email addr from Johannes
% * Explain why we do not do eager interruption (limit sources of interruptions)
% * Finish Non-Interruption section
% * Completely redo the Magritte section
% * Bibliography
% * Pretty Syntax

@@

% We explore the implementation and design challenges that come with such a model, and also show that this model is able to easily create and compose concurrent algorithms.
% The UNIX shell programming model has been shown to be a successful model for composing programs - spawning unrelated programs in parallel to communicate over operating-system pipes.

@@

Our overarching goal is to create a language and a programming system that retains the pipe-based programming and interaction model of bash, but allows for large programs in a way that existing shell languages do not.

\section{Design Considerations}
In this section we discuss several feature requirements and design considerations for a pipe-based language to be viable for large programs.

\paragraph{Not Everything Is a String.}
The first core issue that prevents most shells from being viable for large programs is that only strings are allowed as values.

Several projects, including bash itself, have attempted to add lists and hash-maps as standard objects. Unfortunately, while they allow these values to be stored in variables, most projects do not allow these data structures to be passed into or returned from functions normally.

The \verb/es/ shell (TODO: cite) attempts to extend the standard shell model with lambda functions with lexical scoping. However,

\paragraph{Transactional Channels.}
Since UNIX pipes can only manage bytes, usually channels with a fixed buffer are specified for performance reasons. However, in the languages we consider, we assume we can transactionally write a single complex value to a channel, resulting in the ability to send data in batch without resorting to a queue that may interleave writes.

Consider the fairly common architecture of producer/consumer: A \emph{producer} process producese values that are processed in parallel by multiple \emph{consumer} processes, and then the values are collected in a single output. This might be expressed in bash as:

\begin{lstlisting}
# pass through the stream, labeling every line
label() {
  while read line; do echo "$1$line"; done
}

# process the stream with different labels
split() {
  label a &
  label b &
}

# process the numbers 1-100, limiting the
# output to the first 3 lines
seq 100 | split | head -3
\end{lstlisting}

Unfortunately, when we run this process, the outputs from the \verb/label/ function become interleaved, resulting in outputs such as:

\begin{verbatim}
a1    b12   a
a2    a     b12
a3    a4    a34
\end{verbatim}
In a language intended for large programs, therefore, writes to a channel must be \emph{transactional} - a single write must be sent wholesale to a single reader without any interleaving, no matter how big the write.

\paragraph{Interruption.}
A core requirement of pipe-based languages is a semantic of process \emph{interruption}: automatic clean-up of processes that will no longer be used. Consider the following pseudo-code for a pipe-based language:

\begin{lstlisting}
  read-lines tmp/large-file
    | take 10
    | each (?line => do-expensive-work <@\$@>line)
\end{lstlisting}

In this example, three processes are spawned: (1) a process with a file open that writes one line at a time to its output, (2) a process that reads 10 times from its input, writes each entry to the output, and then exits, and (3) a process that reads every input and calls a function to perform an expensive task.
A user's intent when typing such code may be to read 10 lines from a file and synchronously perform an action on each line.

@@

\paragraph{Compensation.} In order to observe interruption, user code must be able to react to and handle interruption. The most direct way to observe interruption is through a system of \emph{compensation actions}, which are expressions which will be evaluated in the case of an interruption.

For example, the function \verb/file-lines/ above could be implemented as:

\begin{verbatim}
(file-lines ?fname) = (
  handle = (open-file $fname) %%! close-file $fname
  until (=> eof? $handle) (=> read-until "\n" $handle)
)
\end{verbatim}

In this code, the \verb/%%!/ symbol registers a compensation action that will be run either at the next checkpoint (the end of the function), or when the process is interrupted. In this way, we can ensure that the file is closed when the process exits. Compensations are the only way in which interruptions can be directly observed by client code - without compensations, the only way of observing forever-blocked processes would be to inspect the process table, or observe the memory footprint of the program. Thus, we can define the primary task of our system as \emph{running a process's compensation actions at the appropriate time}.

@@

Consider the following example, with the assumption that \\ \verb/do-other-operation/ does not write to the standard output:

@@

\paragraph{Non-closing of Channels.}

@@

\paragraph{Masking Interruptions.} (TODO: explain \verb/produce/ and \verb/consume/)

@@

\section{Description of Magritte.}
